Tail associations in ecological variables and their impact on extinction risk

### 2020-01-24 MainText.Rmd
cleaned upto Methods

<!--ECEs probably produce tail associations-->
There are reasons to hypothesize that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. The idea, here, is based on the reasonable assumption 
that ECEs are both more intense and spatially more extensive than non-extreme 
climatic events. For clarity, we explain the idea using the example case
of extreme rain events. Suppose a single extreme rain event is associated
with spatially widespread large precipitation values, i.e., not only does it rain a lot
in each locale, but it does so for many locales across a large area. Also suppose that smaller local
precipitation measurements are associated, instead, with local, non-extreme weather 
which can differ from location to location, rather than with a single extreme rain event.
The result could be stronger right-tail association, through time, between measurements of precipitation
made in different locations: large precipation measurements are coherent across space, since 
they are all attributable to a single event; whereas small values are less spatially correlated
because they come from a multiplicity of distinct local events.

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$, divided by $T+1$, and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$, divided by $T+1$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. The variance, $\var$,
here and throughout, is the sample variance, with Bessel's correction (denominator $T-1$).
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the normalized ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables. These ideas are revisited in the Discussion.

@@

We simulated our metapopulation models using environmental noises $\epsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variance), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variance), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\epsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cumulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Spearman correlation between $\epsilon_i(t)$ and $\epsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$; likewise partial Spearman correlations were the same. 
The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of the six density-dependent stochastic metapopulation models we used. The population growth rate in habitat patch $i$ at time $t$, $\\lambda_i(t)$ (second column), is $P_i(t+1)/P_i(t)$, and defines how the model works within each patch. Here $P_i(t)$ is the population in patch $i$ at time $t$. The growth rate depends on model parameters, $P_i(t)$, and the environment $\\epsilon_i(t)$ in patch $i$ at time $t$. $\\epsilon_i(t)$ was normally distributed with mean $0$ and standard deviation $\\sigma$, and values were generated independently across times. $P_e$ is the equilibrium of the deterministic, one-patch version of the model. $r_{min}$ is the smallest $r$, given fixed values of the other parameters, for which $P_e$ was non-negative. $r_c$ is the value of $r$ for which a transition occurs from under- to overcompensatory dynamics (see main text). $r_{bf}$ is the value of $r$ for which a transition occurs from a stable to an unstable equilbrium (a bifurcation; see main text). For extinction-risk simulations, we used $r$ such that $r_{min}<r<r_{bf}$. Dispersal between patches was also implemented (see main text). Formulas are given for $P_e$ but specific values which pertain for the listed parameters are given for $r_{min}$, $r_c$ and $r_{bf}$. See Appendix \\ref{SM-model_stability} for the formulas that gave these values.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6,7),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```

>>>>

<!--ECEs probably produce tail associations-->
There are reasons to hypothesize that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. The idea, here, is based on the reasonable assumption 
that ECEs are both more intense and spatially more extensive than non-extreme 
climatic events. For clarity, we explain the idea using the example case
of extreme rain events. Suppose a single extreme rain event is associated
with spatially widespread large precipitation values, i.e., not only does it rain a lot
in each locale, but it does so for many locales across a large area. Also suppose that smaller local
precipitation measurements are associated, instead, with local, non-extreme weather 
which can differ from location to location, rather than with a single extreme rain event.
The result could be stronger right-tail association, through time, between measurements of precipitation
made in different locations: large precipitation measurements are coherent across space, since 
they are all attributable to a single event; whereas small values are less spatially correlated
because they come from a multiplicity of distinct local events.

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$, divided by $T+1$, and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | t=1,\ldots,T\}$, divided by $T+1$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. The variance, $\var$,
here and throughout, is the sample variance, with Bessel's correction (denominator $T-1$).
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the normalized ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables. These ideas are revisited in the Discussion.

@@

We simulated our metapopulation models using environmental noises $\varepsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variance), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variance), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\varepsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cumulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Spearman correlation between $\varepsilon_i(t)$ and $\varepsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$; likewise partial Spearman correlations were the same. 
The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of the six density-dependent stochastic metapopulation models we used. The population growth rate in habitat patch $i$ at time $t$, $\\lambda_i(t)$ (second column), is $P_i(t+1)/P_i(t)$, and defines how the model works within each patch. Here $P_i(t)$ is the population in patch $i$ at time $t$. The growth rate depends on model parameters, $P_i(t)$, and the environment $\\varepsilon_i(t)$ in patch $i$ at time $t$. $\\varepsilon_i(t)$ was normally distributed with mean $0$ and standard deviation $\\sigma$, and values were generated independently across times. $P_e$ is the equilibrium of the deterministic, one-patch version of the model. $r_{min}$ is the smallest $r$, given fixed values of the other parameters, for which $P_e$ was non-negative. $r_c$ is the value of $r$ for which a transition occurs from under- to overcompensatory dynamics (see main text). $r_{bf}$ is the value of $r$ for which a transition occurs from a stable to an unstable equilbrium (a bifurcation; see main text). For extinction-risk simulations, we used $r$ such that $r_{min}<r<r_{bf}$. Dispersal between patches was also implemented (see main text). Formulas are given for $P_e$ but specific values which pertain for the listed parameters are given for $r_{min}$, $r_c$ and $r_{bf}$. See Appendix \\ref{SM-model_stability} for the formulas that gave these values.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6,7),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```
